---
title: "Changes in the demographics of Toronto's shelter system (2018-2024)"
subtitle: "My subtitle if needed"
author: 
  - Dennis Netchitailo
thanks: "Code and data are available at: https://github.com/dennisnetchitailo/Toronto-Shelter-Demographic-Changes"
date: September 27, 2024
date-format: long
abstract: "This paper examines the trends relating to demographic changes in Toronto's shelter system from 2018 to 2024, using publically-accesible data from OpenDataToronto."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(opendatatoronto)
library(knitr)
library(dplyr)
library(ggplot2)
library(readr)
library(lubridate)
analysis_data <- read_csv("data/analysis_data/analysis_data.csv")

```


# Introduction

Since 2018, the number of people using Toronto's shelter system has increased from eight thousand to over ten thousand, an increase of over 25%. For a city facing budgetary constraints, providing shelter to thousands more people than it did half a decade ago is expensive, and leads to the question "Why?". This paper looks at the changes in the shelter's demographic makeup, to try to understand what kind of people having been needing more or less shelter support relative to 2018. 


# Data {#sec-data}

## Raw Dataset {#sec-intro-dataset} 
This paper uses the dataset "Shelter System Flow Data" from OpenDataToronto, which provides data on the types of people in shelters, their ages, and their genders. The time period covered by this dataset is January 2018 to August 2024. This data is collected by the use of the Shelter Management Information System (SMIS) to intake people into various types of shelters (emergency, overnight, hotel/motel programs, etc). 

The limitation of this dataset is that it excludes people not captured by SMIS. This includes individuals who stay outdoors on their own, or who use private & public homeless services that do not track data through SMIS. 
## Cleaned Dataset {#sec-intro-dataset} 
Some of the variables were not assessed and were removed. 
Figure *1* shows a sample of the cleaned dataset.

```{r}
#| label: fig-timeline
#| fig-cap: Trends in the Type of Shelter Users 
#| echo: false
#| warning: false
#| message: false
#| fig-width: 8


print(analysis_data)

```

## Variables in the cleaned Dataset {#sec-variables}
1. **Date**: Records the month & year for which data is provided. The range   
is from January 2018 to August 2024. 
2. **Population**: Records the type of population: either the whole population 
or subgroups that exist within the whole. 
3. **Age under 16**: Records the number of people in the shelter system within 
this age group. 
4. **Age 16-24**: Records the number of people in the shelter system within 
this age group. 
5. **Age 25-34**: Records the number of people in the shelter system within 
this age group. 
6. **Age 35-44**: Records the number of people in the shelter system within 
this age group. 
7. **Age 45-54**: Records the number of people in the shelter system within 
this age group. 
8. **Age 55-65**: Records the number of people in the shelter system within 
this age group. 
9. **Age 65 and over**: Records the number of people in the shelter system within 
this age group. 
10. **Male**: Records the number of people in the shelter system who are male. 
11. **Female**: Records the number of people in the shelter system who are female.
12. **Transgender, other**: Records the number of people in the shelter system who are transgender, non-binary, or two-spirit. 
13. **Population Percentage for a Group**: Records the proportion of the total shelter population that a group occupies. 


### Change in Shelter Subpopulations {#sec-timeline}
```{r}
#| label: fig-timeline
#| fig-cap: Trends in the Type of Shelter Users 
#| echo: false
#| warning: false
#| message: false
#| fig-width: 8

# Convert date format
analysis_data <- analysis_data %>%
  mutate(date_dmy = ymd(date_dmy))

#Filter population groups
selected_groups <- c("Chronic", "Refugees", "Families", "Youth", "Single Adult", "Non-refugees")
filtered_groups <- analysis_data %>%
  filter(population_group %in% selected_groups)

# Create the line plot for the population size 
ggplot(filtered_groups, aes(x = date_dmy, y = gender_male + gender_female + 
                          gender_transgender_non_binary_or_two_spirit, 
                          color = population_group)) +
  geom_line(size = 1) +
  labs(title = "Month-to-Month Change in Population by Subpopulation",
       x = "Date",
       y = "Total Population",
       color = "Population Group") +
  theme_minimal() +
  theme(legend.position = "right")
```

Immediately, it is visible that Youth is the most stable group in terms of population. Families have also not undergone serious change, although why it dipped in 2021 is a question to further look into. Single Adults as a group have been rising, but not unsurprisingly, given the economic challenges Torontonians have been facing since Covid. The group with the wildest swings has been Refugees and Chronic. Both subgroups have more than tripled in number. The number of refugees, also during Covid, plunged, before rapidly growing to a number larger than even before the pandemic. 


# Discussion

## First discussion point {#sec-first-point}


## Second discussion point

## Third discussion point

## Weaknesses and next steps


\newpage

\appendix

# Appendix {-}


# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

\newpage


# References


